---
title: "Let's Plot 8: (Animated) US State Covid-19 Case Count"
author: David McGaughey
date: '2020-03-25'
slug: lets-plot-covid19-animated
categories:
  - R
  - Let's Plot
tags:
  - R
  - Let's Plot
header:
  caption: ''
  image: ''
output:
  blogdown::html_page:
    toc: true
---

## Load packages, pull data
```{r message = FALSE, warning = FALSE, results = FALSE}
library(tidyverse)
library(gganimate)
library(ggrepel)
library(pals)
covid_time <- read_csv('https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv')

covid_death <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Deaths.csv")
```

## Write function to clean and transform
Read the `#comments` for explanations of what is happening

The data has a few inconsistencies in how states/territories are named. 

It's also "wide" and needs to be tidy ("long") for plotting in `ggplot`.
```{r}

# cleaning time
# let's make this a function as we need to do it twice
CSSEGIS_cleaner <- function(table, type = 'Cases'){
  # make state_table to convert two letter states (eg IA) to full name (eg Iowa)
  state_table <- cbind(state.abb, state.name) %>% as_tibble()
  
  out <- table %>% 
    pivot_longer(cols = contains('/20'), names_to = 'Date', values_to  = 'type') %>% 
    # only keep USA, remove 0 case lines
    filter(`Country/Region` == 'US', type > 0) %>% 
    # extract two letter abbreviation, if used (check for comma)
    mutate(State = case_when(grepl(',', `Province/State`) ~ str_extract(`Province/State`, ',.*') %>% 
                               gsub('\\.| |,','',.) %>% 
                               toupper(),
                             TRUE ~ `Province/State`)) %>% 
    # join abbreviation to state_table we made
    left_join(., state_table, by = c('State' = 'state.abb')) %>% 
    # swap in full name into State column
    mutate(State = case_when(is.na(state.name) ~ State,
                             TRUE ~ state.name)) %>% 
    # fix two issues by "hand"
    mutate(State = case_when(State == 'DC' ~ 'District of Columbia',
                             State == 'United States Virgin Islands' ~ 'Virgin Islands',
                             TRUE ~ State)) %>% 
    # remove the cruise ship
    filter(State != 'Grand Princess') %>% 
    # crucial operation where we take the "Dups" (e.g. CA to Californai) and merge them together
    group_by(State, Date) %>% 
    summarise( type = sum( type )) %>% 
    # make date a real date
    mutate(Date = as.Date(Date,format='%m/%d/%y'))
  colnames(out)[3] <- type
  out
}

# run functions
us_covid_cases_clean <- CSSEGIS_cleaner(covid_time)
us_covid_deaths_clean <- CSSEGIS_cleaner(covid_death, 'Deaths')
```

## Plotter function
This is a fairly complex function as it:

  - overlays color lines by `top_states`, given by user, on top of all data (in gray)
  - labels the color lines with `ggrepel` (but only at the end!)
  - uses a `cowplot` theme, with tweaked grid lines
  - custom x and y labels
  - expand x lim to fit the state names
```{r}

covid19_plotter <- function(clean_data, 
                            type = Cases, 
                            top_states = c('Washington', 'New York', 'California'), 
                            xlab = '', 
                            ylab = '', 
                            title = ''){
  plot <- clean_data %>% 
    filter(State %in% top_states) %>% 
    ggplot(aes(x=Date,y= {{ type }}, 
               group = State, 
               color = State)) + 
    geom_line(data = clean_data, aes(x=Date, y = {{ type }}, group = State), color = 'gray') +
    geom_line(size = 2) +
    ggrepel::geom_text_repel(data = clean_data %>% 
                               filter(State %in% top_states) %>% 
                               filter(Date == max(Date)),
                             aes(x=Date, y = {{ type }}, label = State),
                             direction = 'y',
                             nudge_x = 3,
                             hjust = 0,
                             segment.size = 0.5,
                             force = 2,
                             size = 5) +
    cowplot::theme_cowplot() +
    guides(color=FALSE) +
    scale_color_manual(values = as.vector(pals::glasbey(26))) +
    theme(panel.grid.major.y = element_line(linetype = 'dotted')) +
    scale_y_continuous(breaks = c(10, 100,1000,5000,10000, 20000)) +
    scale_x_date(date_labels = "%b %d",
                 breaks = c(as.Date('2020-02-01'),
                            as.Date('2020-02-15'),
                            as.Date('2020-03-01'),
                            as.Date('2020-03-15'),
                            max(clean_data$Date)
                 )) +
    expand_limits(x = c(min(clean_data$Date),max(clean_data$Date) + 7 )) +
    ylab(ylab) + xlab(xlab) + ggtitle(title)
  plot
}
```

## Cases by state 
Important: We don't know the denominator (number of tests given). So New York *may* not have far more cases than the rest of the other states, but rather they are testing far more. 

I've overwritten the function-specified y scale (hence the ggplot warning). 
```{r}
# find top n (by sum cases) states
top_n <- us_covid_cases_clean %>% filter(Date == max(Date)) %>% arrange(-Cases) %>% head(9) %>% pull(State)

covid19_plotter(us_covid_cases_clean, 
                type = Cases,
                top_states = top_n,
                ylab = 'Cases',
                xlab = paste0('\nData Pulled ', Sys.Date(), ' from Johns Hopkins CSSE'),
                title = 'USA COVID-19 Cases by State') +
  scale_y_continuous(breaks = c(100,1000,5000,10000, 20000))
```

### Cases, with log10 scaling
Log10 scaling helps see the trends between states better
```{r}
covid19_plotter(us_covid_cases_clean, 
                type = Cases,
                top_states = top_n,
                ylab = 'log10 Cases',
                xlab = paste0('\nData Pulled ', Sys.Date(), ' from Johns Hopkins CSSE'),
                title = 'USA COVID-19 Cases by State') +
  coord_trans(y = 'log10')
```

## Deaths by state (log10 scaled)
Deaths may be a more useful metric as states may not be testing at comparable rates
```{r}
# find top n (by sum cases) states
top_n <- us_covid_deaths_clean %>% filter(Date == max(Date)) %>% arrange(-Deaths) %>% head(9) %>% pull(State)
covid19_plotter(us_covid_deaths_clean, 
                type = Deaths,
                top_states = top_n,
                ylab = 'log10 Deaths',
                xlab = paste0('\nData Pulled ', Sys.Date(), ' from Johns Hopkins CSSE'),
                title = 'COVID-19 Deaths by US State') +
  coord_trans(y = 'log10')
```

## Deaths by state, animated
This is *extremely* easy to do, thanks to the `gganimate` [package](https://gganimate.com) (written by Thomas Lin Pedersen)
```{r message = FALSE, warning = FALSE}
animation <- covid19_plotter(us_covid_deaths_clean, 
                             type = Deaths,
                             top_states = top_n,
                             ylab = 'log10 Deaths',
                             xlab = paste0('\nData Pulled ', Sys.Date(), ' from Johns Hopkins CSSE'),
                             title = 'COVID-19 Deaths by US State') + 
  coord_trans(y = 'log10') +
  transition_reveal(Date) + 
  view_follow(fixed_y = TRUE, 
              fixed_x = TRUE)

animate(animation, 
        height = 400, 
        width = 600, 
        end_pause = 30)

```

## Shift plot
Many visualizations (e.g. NYTimes, Financial Times) are plotting (on the x-axis) *Days Past n Deaths* so you can better compare the infection rate across different locations.

Let's give that a try.

We need to find, for each state, the first `Date` where `Deaths` >= 5.

We see the state of Washington was about two weeks ahead of everyone else
```{r}
day0 <- us_covid_deaths_clean %>% 
  filter(Deaths >= 5) %>% 
  group_by(State) %>% 
  summarise(Day0 = min(Date))
day0 %>% arrange(Day0)
```
## Transform Data and plot
We filter the data to only retain States with >= 5 deaths

Then we left join the `day0` table we made above

We do math (yes, R can do math on Dates), creating a new `Date` where `Date` = `Date` - `Day0` (`Day0` is first Date in state where COVID19 deaths >= 5).

As our new Date is a time, we swap out the x axis scale for `scale_x_continuous`

We see that New York, Georgia, Lousiana, and New Jersey are on similar trajectories.
```{r}
covid19_plotter(us_covid_deaths_clean %>% 
                  filter(State %in% day0$State,
                         Deaths >= 5) %>% 
                  left_join(., day0, by = 'State') %>% 
                  mutate(Date = Date - Day0), 
                type = Deaths,
                top_states = top_n,
                ylab = 'log10 Deaths',
                xlab = paste0('Days Past 5 Deaths\n\nData Pulled ', Sys.Date(), ' from Johns Hopkins CSSE'),
                title = 'COVID-19 Deaths by US State') +
  scale_x_continuous() + 
  coord_trans(y = 'log10') +
  scale_y_continuous(breaks = c(10,25, 50,75, 100))
```
