---
title: "Let's Plot 9: The venerable box plot"
author: David McGaughey
date: '2020-07-23'
slug: lets-plot-box-plot-fancy
categories:
  - R
  - Let's Plot
tags:
  - R
  - Let's Plot
header:
  caption: ''
  image: ''
output:
  blogdown::html_page:
    toc: true
---
## Intro
This is the 9th Let's Plot...and I've not done a workup of the *most* useful plot - the boxplot. Oops. Well let's rectify that. 

## Load packages

Many many packages. We'll see what they do later.

If you get an error loading the packge, you can install it by running `install.packages` as:
`install.package('dabestr')`
```{r}
library(dabestr)
library(readr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggbeeswarm)
library(cowplot)
```

## Import Excel file

Quick peek into the import. This data is from Figure 2a from the [Bharti lab](https://stm.sciencemag.org/content/11/475/eaat5580?rss=1)
```{r}
fig2a_data <- read_tsv('https://github.com/davemcg/davemcg.github.io/raw/master/content/post/fig2.tsv')

fig2a_data
```

This doesn't look right...there's no header in this data. Let's re-import.
```{r}
fig2a_data <- read_tsv('https://github.com/davemcg/davemcg.github.io/raw/master/content/post/fig2.tsv', col_names = FALSE)
fig2a_data
```

Better. Let's add some column names

```{r}
colnames(fig2a_data) <- c('Scaffold', 'Young\'s Modulus')
fig2a_data
```

OK, let's see how many `Scaffold` types we have and the number of samples of each with `base R` style.

```{r}
table(fig2a_data$Scaffold )
```

Now let's do the same, but in a tidyverse way. We us the ` %>% ` to *pipe* the `fig2a_data` into `group_by` then the grouped data is sent to `summarise` to do the counting (with `n()`). It's more readable this way.

```{r}
fig2a_data %>% group_by(Scaffold) %>% summarise(Count = n())
```
## Plotting!
OK. Let's do a quick boxplot. 

```{r}
fig2a_data %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot()
```

## Hmm, the order is not ideal
```{r}
fig2a_data %>% 
  mutate(Scaffold = factor(Scaffold, levels = c('Single layer unfused', 'Single layer fused', 'Bilayer random fibers', 'Bilayer aligned fibers'))) %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot()
```

## Overlay points

I am a HUGE proponent of showing your data. If you *just* use the boxplot you are not giving any information on how many experiments you have done. 

So let's show off the data by using the the *modular* nature of ggplot

Need another plotting style? Just `+` another `geom_`

```{r}
fig2a_data %>% 
  mutate(Scaffold = factor(Scaffold, levels = c('Single layer unfused', 'Single layer fused', 'Bilayer random fibers', 'Bilayer aligned fibers'))) %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot() +
  geom_point()
```

## Wilcox test

It's popular in science papers to show the p-values between the groups. We set up the groupings with the `my_comparison` list object. `ggpubr` then is used to overlay the results with its `stat_compare_means` ggplot add-on.
```{r}
my_comparisons <- list( c("Single layer unfused", "Single layer fused"), c("Single layer unfused", "Bilayer random fibers"), c("Single layer unfused", "Bilayer aligned fibers") )
fig2a_data %>% 
  mutate(Scaffold = factor(Scaffold, levels = c('Single layer unfused', 'Single layer fused', 'Bilayer random fibers', 'Bilayer aligned fibers'))) %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot() +
  geom_point() +
  stat_compare_means(comparisons = my_comparisons, method = 'wilcox.test')
```

## ggbeeswarm
When you have many points, they tend to overlap each other. A *beeswarm* plot is nice because it spreads the points sideways in a manner to show the distribution of the data.

I've replaced `geom_point` with `geom_quasirandom`.

```{r}
my_comparisons <- list( c("Single layer unfused", "Single layer fused"), c("Single layer unfused", "Bilayer random fibers"), c("Single layer unfused", "Bilayer aligned fibers") )
fig2a_data %>% 
  mutate(Scaffold = factor(Scaffold, levels = c('Single layer unfused', 'Single layer fused', 'Bilayer random fibers', 'Bilayer aligned fibers'))) %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot() +
  geom_quasirandom() +
  stat_compare_means(comparisons = my_comparisons, method = 'wilcox.test') 
```

## Themes

I'm not partial to gray everwhere, so let's use the `cowplot` `theme_cowplot` theme. 
```{r}
my_comparisons <- list( c("Single layer unfused", "Single layer fused"), c("Single layer unfused", "Bilayer random fibers"), c("Single layer unfused", "Bilayer aligned fibers") )
fig2a_data %>% 
  mutate(Scaffold = factor(Scaffold, levels = c('Single layer unfused', 'Single layer fused', 'Bilayer random fibers', 'Bilayer aligned fibers'))) %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot() +
  geom_quasirandom() +
  stat_compare_means(comparisons = my_comparisons, method = 'wilcox.test') +
  theme_cowplot() 
```

## Themes, with some tweaking of color and text

  - Increase point size
  - Make boxplot black
  - Make points gray 
  - Make scaffold labels less wide by adding `\n` (newline) between each word
  
```{r}
my_comparisons <- list( c("Single\nlayer\nunfused", "Single\nlayer\nfused"), c("Single\nlayer\nunfused", "Bilayer\nrandom\nfibers"), c("Single\nlayer\nunfused", "Bilayer\naligned\nfibers") )
fig2a_data %>% 
  mutate(Scaffold = gsub(' ', '\n', Scaffold), 
         Scaffold = factor(Scaffold, levels = c('Single\nlayer\nunfused', 'Single\nlayer\nfused', 'Bilayer\nrandom\nfibers', 'Bilayer\naligned\nfibers'))) %>% 
  ggplot(aes(x=Scaffold, y=`Young\'s Modulus`)) +
  geom_boxplot(fill = 'black', color = 'dimgray') +
  geom_quasirandom(color = 'dimgray', size = 3) +
  stat_compare_means(comparisons = my_comparisons, method = 'wilcox.test') +
  theme_cowplot() 
```


## dabest, one comparison
One visualization that has been discussed as an alternative to the boxplot is the [estimation plot](https://www.nature.com/articles/s41592-019-0470-3). As R is a super popular figure making platform, the authors have made an [R package](https://cran.r-project.org/web/packages/dabestr/vignettes/using-dabestr.html
) to make these plots. 




```{r}
dabest_dat <- fig2a_data %>% 
  filter(Scaffold %in% c("Single layer unfused", "Single layer fused"))
dabest(dabest_dat, x= Scaffold, y = `Young\'s Modulus`,
                             idx = c("Single layer unfused", "Single layer fused"),
                             paired = FALSE) %>% 
                      mean_diff() %>% 
  plot(effsize.ylim = c(0, 1000))
```

## dabest, multiple comparisons
```{r}
dabest(fig2a_data, x= Scaffold, y = `Young\'s Modulus`,
                             idx = c("Single layer unfused", "Single layer fused", "Bilayer random fibers","Bilayer aligned fibers"),
                             paired = FALSE) %>% 
                      mean_diff() %>% 
  plot(effsize.ylim = c(0, 1000))
```

## Conclusion

Boxplots are easy, making them look *awesome* is more work. But R and ggplot2 are up to it. 

## Session Info 
Version numbers of things
```{r}
sessionInfo()
```